Tesseroids: forward modeling gravitational fields in spherical coordinates

### 2015-03-07 manuscript.tex
Finish first try at the conclusions

>>>>

We have presented the open-source software \textit{Tesseroids}.
It consists of command-line programs,
written in the C programming language,
to perform the forward modeling of gravitational fields in spherical
coordinates.
The fields are calculated from a mass model composed of spherical prisms, the
so-called tesseroids.
The volume integrals of the gravitational fields of a tesseroid are solved
numerically using the Gauss-Legendre Quadrature (GLQ).
The GLQ approximates the volume integrals by weighted sums of point mass
effects.
The error of the GLQ integration increases as the computation point gets closer
to the tesseroid.
To counter this effect,
the accuracy of the GLQ integration can be increased by using more point
masses or by dividing each tesseroid into smaller ones.

We have implemented and improved upon an adaptive discretization algorithm to
achieve an optimal division of tesseroids.
Tesseroids are divided into more parts closer to the computation point,
where more point masses are needed.
Our implementation of the adaptive discretization uses a ``stack'' data
structure in place of the originally proposed recursive implementation.
The stack-based algorithm is computationally faster and allows more control
over errors when too many divisions are necessary.
The adaptive discretization is controlled by
a scalar called the distance-size ratio ($D$).
The algorithm ensures that all tesseroids will
have dimensions smaller than $D$ times the distance to the computation point.
The value of $D$ indirectly controls the accuracy of the integration as well as
the computation time.

We performed an error analysis to determine
the optimal value of $D$ required to achieve a target accuracy.
We used a spherical shell as a reference to calculate the computation error of
our algorithm for different values of $D$.
Our results show that the values of $D$ required to achieve a maximum error
of 0.1\% of the shell values are
1 for the gravitational potential, 1.5 for the gravitational acceleration,
and 8 for the gravity gradients.
Previous assumptions were that accurate results are guaranteed if
the distance to the tesseroid is larger than
the distance between point masses.
This same condition was previously applied indiscriminately
to both the gravitational acceleration and the gravity gradients.
That assumption is equivalent to using $D=1.5$ for all fields.
Our results show that this is valid for the gravitational
acceleration and results in a 0.1\% computation error.
This is expected because the original study that determined the above condition
was performed on the vertical component of gravitational acceleration.
However, applying the same condition to the gravity gradients produces
an error of the order of $10^2\%$.

For the gravity gradients in particular,
the distance-size ratio required for 0.1\% error decreases with height.
This can be attributed to the higher sensitivity of gravity gradients
to height variations because they decay with the distance cubed.
This behavior causes the integration kernel to have more high frequency
components, thus requiring more point masses for numerical integration.

The values of the distance-size ratio determined above were
incorporated as defaults in the software \textit{Tesseroids}.
We chose the value $D=8$ for the gravity gradients as a conservative default.
If the user desires, the value of $D$ used can be controlled by a command-line
argument.

In situations that require many tesseroid divisions,
the stack used in the algorithm will overflow and further
divisions become impossible.
The current implementation warns the user that
the overflow occurred and proceeds with the GLQ integration without division.
Future improvements to the algorithm include a better way to handle such
situations as they arise.
An alternative would be to replace the tesseroid by an equivalent
right rectangular prism and compute its effects instead.
This would allow accurate computations at smaller distances.
Furthermore,
the computation time increases drastically as the computation point
gets closer to the tesseroid.
This effect can be prohibitive for computing the gravity gradients at
relatively low heights (e.g., for terrain corrections of
ground or airborne surveys).
Further investigation of different criteria for dividing the tesseroids could
yield better performance through a reduced number of divisions.
